It's all about Meaning!\\{\normalsize A Theory for Data Migration when Deploying Changes

### 2022-12-18 2022 Migration/articleMigration.tex
fixed subst of concepts in inst
rewritten the abstract

\title{It's all about Meaning!\\{\normalsize A Theory for Data Migration during Software Development}}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\author[umn]{Sebastiaan Joosten\fnref{fn2}}
\address[ou]{Open Universiteit Nederland, Heerlen, the Netherlands}
\address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
\address[umn]{University of Minnesota, Minneapolis, USA}
\fntext[fn1]{ORCID 0000-0001-8308-0189}
\fntext[fn2]{ORCID 0000-0002-6590-6220}

\begin{abstract}
   The Ampersand project has provided the theory and tools to generate information systems from an algebraic specification.
   However, information systems in practice may change repeatedly after their maiden deployment.
   Changes that affect the data model typically result in a data migration.
   In such cases, simply regenerating the system is not enough.
   That would reset the database to its initial state, losing all data gathered so far.
   To prevent that, a migration engineer must transfer the old data to the new system by hand.

   In this contribution we develop a theory for reliable data migration to help the migration engineer
   to transfer the data and preserve their semantics as much as possible.
   We aim to automate the data migration,
   to prevent mistakes and enable more frequent migrations.
   The target is to generate a migration script from two specifications: the old specification and the new specification.
   A software generator that embodies this theory is subject of future research.
\end{abstract}

\begin{keyword}
relation algebra\sep software development\sep data migration\sep software migration\sep Ampersand
\end{keyword}
\maketitle

@@

\subsection{Data migration}
   In a datamigration, we distinguish three moments:
\begin{enumerate}
   \item the moment of transition, i.e. the moment the old system is replaced (in production) by the new system;
   \item the decision point, i.e. the moment a business authority declares the migration successful;
   \item the point of no return, i.e. the moment that the migration cannot be undone other than by the regular backup-procedure.
\end{enumerate}
   The decision point must come after the moment of transition, to allow the business to verify the migration.
   The point of no return must come after the decision point, to allow for a no-go decision from the business.

   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
\begin{enumerate}
   \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and the functionality\footnote{For now, we ignore the functionality.}.
         In Ampersand, this means that the developer produces an Ampersand script for the new system.
         The difference between the new system and the old system is the increment.
         To preserve the semantics of data means that the dataset $\dataset$ must be combined with $\dataset'$ in the new system.
         Dataset $\dataset'$ is optional, in case the developer wants to add some new data into the system on the moment of transition.
         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\infsys'$,
         which gives the user the full functionality of the old system and the new system,
         i.e.\ $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
         TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
         yielding migration system $\migrsys'$.
         The migration engineer does not alter the old system $\infsys$ nor the specification new system $\infsys'$ that the developer has produced.
         As a result, the migration engineer can always restore the old system because it is part of the migration system.
   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
         while $\dataset_{\infsys}\subseteq\dataset_{\migrsys'}$.
         This means that the data of the old system remains intact.
         This event marks the moment of transition.
   \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
         to enforce their execution.
         Automatic migration rules can be executed at the moment of transition.
         Migration rules that are enforced manually will be held alive until they are satisfied.
         All rules from $\schema$ (the rules from the old system) are removed on the moment of transition, because the are satisfied by property~\ref{eqn:satisfaction}.
   \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
   \item Users may have to perform manual migration steps.
         These are specified as process rules in $\id{migrationrules}$.
         % Dit zijn "kurk" regels.
         The system ensures that the amount of work needed to perform these steps is finite.
   \item The patron of the system can decide whether to carry on with the migration or to roll back to the old system.
         She can do that incrementally by switching off the interfaces one-by-one.
   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
\end{enumerate}

@@

To implement the disjoint union, we need to relabel 
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\{a\ \inst\ \subst{C}{D}{A}\mid a\ \inst\ A\}}
   \end{array}\]
\end{definition}
\begin{definition}[relabel concepts in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
\end{definition}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{t}\mid t\in\triples\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}

>>>>

\title{It's all about Meaning!\\{\normalsize A Theory for Data Migration when Deploying Changes}}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\author[umn]{Sebastiaan Joosten\fnref{fn2}}
\address[ou]{Open Universiteit Nederland, Heerlen, the Netherlands}
\address[ordina]{Ordina NV, Nieuwegein, the Netherlands}
\address[umn]{University of Minnesota, Minneapolis, USA}
\fntext[fn1]{ORCID 0000-0001-8308-0189}
\fntext[fn2]{ORCID 0000-0002-6590-6220}

\begin{abstract}
   To support incremental software development with software generators bears the potential of decreasing the time to deploy changes.
   However, an increment that changes the database schema causes a nontrivial data migration problem.
   The problem is how to preserve the semantics of the data as much as possible and satisfy elementary business requirements at the same time.

   The Ampersand project has provided the theory and tools to generate information systems from an algebraic specification.
   This paper develops a theory for making changes to an already deployed information system that was generated by Ampersand.
   We propose a migration method with zero down-time, and allows the business to undo the change after verification.
   It also leaves room for the business to finish ongoing work after the moment the change is deployed.
   We aim to automate the change including the data migration,
   to prevent mistakes and enable more frequent changes.
\end{abstract}

\begin{keyword}
generative software\sep incremental software development\sep data migration\sep relation algebra \sep Ampersand
\end{keyword}
\maketitle

@@

\subsection{Data migration}
   In a datamigration, we distinguish three moments:
\begin{enumerate}
   \item the moment of transition, i.e. the moment the old system is replaced (in production) by the new system;
   \item the decision point, i.e. the moment a business authority declares the migration successful;
   \item the point of no return, i.e. the moment that the migration cannot be undone other than by a backup-procedure,
         which is outside the scope of this paper.
\end{enumerate}
   The decision point must come after the moment of transition, to allow the business to verify the migration.
   The point of no return must come after the decision point, to allow for a no-go decision from the business.

   To migrate system $\infsys$ to $\infsys'$, we take the following steps:
\begin{enumerate}
   \item Let the old system $\infsys$ have state $\dataset$ and schema $\schema$.
         A developer defines a new system $\infsys'$, which contains a dataset $\dataset'$, a schema $\schema'$, and some functionality\footnote{For now, we ignore the functionality.}.
         In Ampersand, this means that the developer produces an Ampersand script for the new system.
         The difference between the new system and the old system is the increment.
         The new system must contain the triples from $\dataset$ and $\dataset'$, to preserve the old data.
         If the developer wants to add some new data into the system on the moment of transition,
         she specifies that in $\dataset'$.
         That might be necessary to initialize specific new features.
         If $\schema$ equals $\schema'$, we call the migration trivial because a migration is not necessary.
   \item The generator generates a migration system $\migrsys$ based on $\schema$ and $\infsys'$,
         which gives the user the full functionality of the old system and the new system,
         i.e.\ $\schema_{\migrsys}=(\schema\sqcup\schema')\cup\id{migrationrules}$.
         The schema $\id{migrationrules}$ represents the enforce rules that migrate the existing data to the new dataset.
         TODO: $\id{migrationrules}$ bevat alleen tijdelijke datastructuur, indien nodig.
   \item A migration engineer changes $\id{migrationrules}$ to accommodate the non-standard requirements of the migration,
         yielding migration system $\migrsys'$.
         The migration engineer does not alter the old system $\infsys$ nor the specification new system $\infsys'$ that the developer has produced.
         As a result, the migration engineer can always restore the old system because it is part of the migration system.
   \item The production environment switches instantly from $\infsys$ to $\migrsys'$,
         while $\dataset_{\infsys}\subseteq\dataset_{\migrsys'}$.
         This means that the data of the old system remains intact.
         This event marks the moment of transition.
   \item The rules in the migration system that are not in $\schema'$ will remain in the system until they are satisfied,
         to enforce their execution.
         Automatic migration rules can be executed at the moment of transition.
         Migration rules that are enforced manually will be held alive until they are satisfied.
         All rules from $\schema$ (the rules from the old system) are removed on the moment of transition, because the are satisfied by property~\ref{eqn:satisfaction}.
   \item After a finite span of time, $\dataset_{\migrsys'}$ satisfies all its enforce rules.
         This implies that the generated rules plus the changes made by the migration engineer ensure the preservation of the meaning of the old data.
   \item Users may have to perform manual migration steps.
         These are specified as process rules in $\id{migrationrules}$.
         % Dit zijn "kurk" regels.
         The system ensures that the amount of work needed to perform these steps is finite.
   \item The patron of the system can decide whether to carry on with the migration or to roll back to the old system.
         She can do that incrementally by switching off the interfaces one-by-one.
   \item The system retains the data, the schema, and the functionality that is covered by $\schema'$, dropping everything else.
\end{enumerate}

@@

To implement the disjoint union, we need to relabel 
\begin{definition}[relabel concepts]
   \[\subst{C}{D}{A}\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ D\ \text{\bf else}\ A\]
\end{definition}
\begin{definition}[relabel concepts in triples]
   \[\subst{C}{D}{\triple{a}{r}{b}} = \triple{a}{\subst{C}{D}{r}}{b}\]
\end{definition}
\begin{definition}[relabel concepts in relations]
   \[\subst{C}{D}{(\declare{n}{A}{B})} = \declare{n}{\subst{C}{D}{A}}{\subst{C}{D}{B}}\]
\end{definition}
\begin{definition}[relabel concepts in $\inst$]
   \[a\ \subst{C}{D}{\inst}\ A\ =\ \text{\bf if}\ C=A\ \text{\bf then}\ a\ \inst\ D\ \text{\bf else}\ a\ \inst\ A\]
\end{definition}
\begin{definition}[relabel concepts in datasets]
   \[\begin{array}{rcl}
      \rlap{$\subst{C}{D}{\pair{\triples}{\inst}}$}\\
      &=&\pair{\{\subst{C}{D}{t}\mid t\in\triples\}}{\subst{C}{D}{\inst}}
   \end{array}\]
\end{definition}
\begin{definition}[relabel concepts in rules]
   \[\sat{u}{\dataset}\ \Leftrightarrow\ \sat{\subst{C}{D}{u}}{\subst{C}{D}{\dataset}}\]
\end{definition}
\begin{definition}[relabel concepts in schemas]
   \begin{eqnarray}
      \rlap{$\subst{C}{D}{\triple{\concepts}{\rels}{\rules}}$}\notag\\
      &=&\begin{array}[t]{l@{}l}
         \la&\{\subst{C}{D}{t}\mid t\in\triples\}\\
         ,&\{\subst{C}{D}{r}\mid r\in\rels\}\\
         ,&\{\subst{C}{D}{u}\mid u\in\rules\}\ \ra\notag
         \end{array}
   \end{eqnarray}
\end{definition}
